---
title: "Downloading & Processing publicly available datasets"
author: "Victor Banerjee"
date: "March 18, 2026"
output:
  html_document:
    toc: true
    theme: united
---

```{r logger_package_note}
# install.packages("futile.logger")
```

In this vignette, we will download normalized gene expression data from two 
sources - recount2 and CBTN and process them to save as log2 normalized counts.

# Installing Library

```{r}
# BiocManager::install("recount")
```

# Loading Libraries

```{r}
library(recount)
library(futile.logger)
```


# Path configuration
- Set up project directory structure and helper functions.
- Folders Structure:
  - data/raw/recount_brain: raw recount2 files
  - data/raw/openpbta: raw OpenPBTA files
  - data/processed/log2_tpm_rse: processed log2(TPM+1) matrices
  - data/processed/brain_umap: UMAP coordinates and metadata
  - output/final_processed: final combined matrices and metadata for analysis

```{r}
# Biologist note:
# All raw and processed files are kept under this repository's data folder.
# This block creates folders if they do not already exist.

project_dir = normalizePath(".", winslash = "/", mustWork = FALSE)
project_data_dir = file.path(project_dir, "data")
raw_recount_dir = file.path(project_data_dir, "raw", "recount_brain")
raw_openpbta_dir = file.path(project_data_dir, "raw", "openpbta")
processed_dir = file.path(project_data_dir, "processed")
log2_tpm_dir = file.path(processed_dir, "log2_tpm_rse")
brain_umap_dir = file.path(processed_dir, "brain_umap")
output_dir = file.path(project_dir, "output")
final_output_dir = file.path(output_dir, "final_processed")

dir.create(project_data_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(raw_recount_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(raw_openpbta_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(processed_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(log2_tpm_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(brain_umap_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(final_output_dir, recursive = TRUE, showWarnings = FALSE)

# Helper to run functions that require a working directory, without hard-coding setwd in the notebook flow.
run_in_dir = function(path, expr) {
  old_dir = getwd()
  on.exit(setwd(old_dir), add = TRUE)
  dir.create(path, recursive = TRUE, showWarnings = FALSE)
  setwd(path)
  force(expr)
}
```

# downloading data from recount2
- Recount2 provides uniformly processed RNA-seq data from public studies, including TCGA and GTEx. 
- Recount2's R package allows programmatic access to metadata and expression matrices, which we will use to build our combined dataset. 
- Each study's data is stored as a SummarizedExperiment object, which we will process into log2(TPM+1) format and subset to protein-coding genes.



```{r}
# Biologist note:
# This block downloads recount-brain study objects and metadata,
# which provide public RNA-seq sources used later for adult-brain references.

rm(list=ls())
library(recount)

if (!requireNamespace("futile.logger", quietly = TRUE)) {
  stop("Package 'futile.logger' is required. Please run install.packages('futile.logger').")
}
library(futile.logger)

project_dir = normalizePath(".", winslash = "/", mustWork = FALSE)
project_data_dir = file.path(project_dir, "data")
raw_recount_dir = file.path(project_data_dir, "raw", "recount_brain")
dir.create(raw_recount_dir, recursive = TRUE, showWarnings = FALSE)

flog.appender(appender.console(), name = "BRAINUMAP")
flog.threshold(INFO, name = "BRAINUMAP")
log_step = function(msg) {
  futile.logger::flog.info(msg, name = "BRAINUMAP")
  message(msg)
}
log_step("START: Downloading recount metadata and study files")

run_in_dir = function(path, expr) {
  old_dir = getwd()
  on.exit(setwd(old_dir), add = TRUE)
  dir.create(path, recursive = TRUE, showWarnings = FALSE)
  setwd(path)
  force(expr)
}

meta_v2 = recount::add_metadata(source = 'recount_brain_v2')
meta_v1 = recount::add_metadata(source = 'recount_brain_v1')
log_step("Metadata download complete")
length(unique(meta_v1$sra_study_s))
length(unique(meta_v2$sra_study_s))
setdiff( unique(meta_v2$sra_study_s), unique(meta_v1$sra_study_s) ) # gtex study is in v2/
save(meta_v2, file = file.path(raw_recount_dir, "recount_brain_v2.Rdata"))
log_step("Saved metadata to data/raw/recount_brain/recount_brain_v2.Rdata")


studies = na.omit(unique(meta_v2$sra_study_s))
lapply(studies, function(x){
    message(x)
  if(!file.exists(file.path(raw_recount_dir, x, 'rse_gene.Rdata'))) {
    log_step(paste0("Downloading gene-level recount study: ", x))
    run_in_dir(raw_recount_dir, download_study(x))
    }
})

lapply(studies, function(x){
    message(x)
  if(!file.exists(file.path(raw_recount_dir, x, 'rse_tx.Rdata'))) {
    log_step(paste0("Downloading transcript-level recount study: ", x))
    run_in_dir(raw_recount_dir, download_study(x, type="rse-tx"))
    }
})

if(!file.exists(file.path(raw_recount_dir, 'TCGA', 'rse_gene_brain.Rdata'))) {
    
  dir.create(file.path(raw_recount_dir, 'TCGA'), recursive = TRUE, showWarnings = FALSE)
    downloader::download('http://duffel.rail.bio/recount/v2/TCGA/rse_gene_brain.Rdata', destfile = 
               file.path(raw_recount_dir, 'TCGA', 'rse_gene_brain.Rdata'), mode = 'wb')
}
log_step("END: Recount download block finished")
```

# Creating log2 TPM counts. 
- This block converts TCGA gene counts into log2(TPM+1), then keeps protein-coding genes for compatibility with downstream comparisons.
- We also build a GTEx log2(TPM+1) matrix in the same format, to serve as a normal brain reference in our combined dataset.

```{r}
# Biologist note:
# This block converts TCGA gene counts into log2(TPM+1),
# then keeps protein-coding genes for compatibility with downstream comparisons.

rm(list=ls())
library(rtracklayer)
library(recount)

if (!requireNamespace("futile.logger", quietly = TRUE)) {
  stop("Package 'futile.logger' is required. Please run install.packages('futile.logger').")
}
library(futile.logger)

project_dir = normalizePath(".", winslash = "/", mustWork = FALSE)
project_data_dir = file.path(project_dir, "data")
raw_recount_dir = file.path(project_data_dir, "raw", "recount_brain")
resdir = file.path(project_data_dir, "processed", "log2_tpm_rse")
annotations_dir = file.path(raw_recount_dir, "annotations")
gtf_path = file.path(annotations_dir, "gencode.v25.annotation.gtf.gz")
dir.create(resdir, recursive = TRUE, showWarnings = FALSE)
dir.create(annotations_dir, recursive = TRUE, showWarnings = FALSE)

flog.appender(appender.console(), name = "BRAINUMAP")
flog.threshold(INFO, name = "BRAINUMAP")
log_step = function(msg) {
  futile.logger::flog.info(msg, name = "BRAINUMAP")
  message(msg)
}
log_step("START: Building TCGA log2(TPM+1) matrix")

rse_gene = get(load(file.path(raw_recount_dir, "TCGA", 'rse_gene_brain.Rdata')))
if (!file.exists(gtf_path)) {
  log_step("Missing gencode.v25.annotation.gtf.gz - downloading to data/raw/recount_brain/annotations")
  downloader::download(
    url = "https://ftp.ebi.ac.uk/pub/databases/gencode/Gencode_human/release_25/gencode.v25.annotation.gtf.gz",
    destfile = gtf_path,
    mode = "wb"
  )
}
gtf = import(gtf_path)
gtf = gtf[gtf$gene_type=="protein_coding", ]
gtf = gtf[gtf$type=="gene", ]
idx = match(gtf$gene_id, names(rowRanges(rse_gene)))
export(gtf, con= file.path(resdir, "prot_coding_gencode.v25.annotation.gtf"), format = "GTF")

# convert tcga to log2(tpm+1) 
rpkm <- getRPKM(scale_counts(rse_gene))
tpm = apply(rpkm, 2, function(x){
    (x/sum(x))*10^6
})
log2_tpm =log2(tpm+1)
assay(rse_gene) = log2_tpm

# subset to only protein-coding genes. 	
rse_gene = rse_gene[idx, ]
rowRanges(rse_gene)$gencode_gene_name = gtf$gene_name
save(rse_gene, file = file.path(resdir, paste0( "TCGA_rse_gene.RData")) )
log_step("END: Saved processed TCGA matrix to data/processed/log2_tpm_rse/TCGA_rse_gene.RData")

# Build and save GTEX in the same processed format expected downstream.
meta_v2_path = file.path(raw_recount_dir, "recount_brain_v2.Rdata")
if (!file.exists(meta_v2_path)) {
  stop(
    paste0(
      "Missing metadata file: ", meta_v2_path, "\n",
      "Run the recount download block first to create recount_brain_v2.Rdata."
    )
  )
}
meta_v2 = get(load(meta_v2_path))
gtex_studies = unique(na.omit(meta_v2[meta_v2[, "Dataset"] == "GTEX", "sra_study_s"]))
if (length(gtex_studies) == 0) {
  stop("Could not find a GTEX study ID in recount_brain_v2 metadata.")
}
gtex_study = gtex_studies[1]
gtex_source_path = file.path(raw_recount_dir, gtex_study, "rse_gene.Rdata")

if (!file.exists(gtex_source_path)) {
  log_step(paste0("Missing GTEX source file. Attempting download for study: ", gtex_study))
  old_dir = getwd()
  on.exit(setwd(old_dir), add = TRUE)
  setwd(raw_recount_dir)
  download_study(gtex_study)
}

if (!file.exists(gtex_source_path)) {
  stop(
    paste0(
      "Missing GTEX source file after download attempt: ", gtex_source_path, "\n",
      "Check network access and recount availability, then rerun."
    )
  )
}

log_step("START: Building GTEX log2(TPM+1) matrix")
gtex_rse = get(load(gtex_source_path))
rpkm_gtex <- getRPKM(scale_counts(gtex_rse))
tpm_gtex = apply(rpkm_gtex, 2, function(x){
  (x/sum(x))*10^6
})
log2_tpm_gtex = log2(tpm_gtex + 1)
assay(gtex_rse) = log2_tpm_gtex

# Keep identical protein-coding gene set/order as TCGA reference.
gtex_rse = gtex_rse[idx, ]
rowRanges(gtex_rse)$gencode_gene_name = gtf$gene_name
save(gtex_rse, file = file.path(resdir, "GTEX_rse_gene.RData"))
log_step("END: Saved processed GTEX matrix to data/processed/log2_tpm_rse/GTEX_rse_gene.RData")

```


## 
```{r}
library(futile.logger)

project_dir = normalizePath(".", winslash = "/", mustWork = FALSE)
project_data_dir = file.path(project_dir, "data")
raw_recount_dir = file.path(project_data_dir, "raw", "recount_brain")
resdir = file.path(project_data_dir, "processed", "log2_tpm_rse")
```

## CGGA Dataset Processing
- The CGGA dataset (SRP027383) is processed similarly to TCGA and GTEx, but since it is not already in recount2, we start from the raw RSE object and apply the same log2(TPM+1) transformation and gene filtering.

```{r}

library(SummarizedExperiment)
library(recount)
library(rtracklayer)
library(futile.logger)



log_step = function(msg) { futile.logger::flog.info(msg); message(msg) }

# Load the protein-coding gene index from already-processed TCGA object
tcga_rse       = get(load(file.path(resdir, "TCGA_rse_gene.RData")))
target_gene_ids = rownames(tcga_rse)

# Load your SRP027383 rse_gene
rse = get(load(file.path(raw_recount_dir, "SRP027383", "rse_gene.Rdata")))
log_step(paste0("SRP027383 dimensions: ", nrow(rse), " genes x ", ncol(rse), " samples"))

# Normalize to log2(TPM+1)
rpkm_mat = getRPKM(scale_counts(rse))
tpm_mat  = apply(rpkm_mat, 2, function(x) (x / sum(x)) * 1e6)
assay(rse) = log2(tpm_mat + 1)

# Subset to protein-coding genes matching TCGA/GTEX
idx = match(target_gene_ids, rownames(rse))
log_step(paste0("Genes matched: ", sum(!is.na(idx)), " / ", length(target_gene_ids)))

other_brain_rse_gene = rse[na.omit(idx), ]
rownames(other_brain_rse_gene) = target_gene_ids[!is.na(idx)]

save(other_brain_rse_gene,
     file = file.path(resdir, "other_brain_rse_gene_6_3_2020.RData"))
log_step("Saved other_brain_rse_gene_6_3_2020.RData")
```

## Combining data from recount2
- This block combines TCGA, GTEx, and CGGA into one harmonized expression matrix, then appends glioma-related annotations (IDH, 1p/19q, grade, subtype labels).
- The resulting combined matrix will be used for downstream analyses and visualizations, such as UMAP projections colored by gene expression or mutation status.
```{r}
rm(list=ls())
library(SummarizedExperiment)
library(recount)
library(rtracklayer)
library(futile.logger)

project_dir      = normalizePath(".", winslash = "/", mustWork = FALSE)
project_data_dir = file.path(project_dir, "data")
raw_recount_dir  = file.path(project_data_dir, "raw", "recount_brain")
resdir           = file.path(project_data_dir, "processed", "log2_tpm_rse")

log_step = function(msg) { futile.logger::flog.info(msg); message(msg) }

# Load protein-coding gene index
tcga_rse        = get(load(file.path(resdir, "TCGA_rse_gene.RData")))
target_gene_ids = rownames(tcga_rse)

# Load SRP027383
rse = get(load(file.path(raw_recount_dir, "SRP027383", "rse_gene.Rdata")))
log_step(paste0("SRP027383 dimensions: ", nrow(rse), " genes x ", ncol(rse), " samples"))

# Normalize to log2(TPM+1)
rpkm_mat = getRPKM(scale_counts(rse))
tpm_mat  = apply(rpkm_mat, 2, function(x) (x / sum(x)) * 1e6)
assay(rse) = log2(tpm_mat + 1)

# --- Replace colData with meta_v2 via key join ---
# Key: rse$sample (SRS accession per column) <-> meta_v2$sra_sample_s
meta_v2 = get(load(file.path(raw_recount_dir, "recount_brain_v2.Rdata")))

rse_samples  = as.character(rse$sample)          # one value per column
meta_idx     = match(rse_samples, meta_v2$sra_sample_s)

log_step(paste0("Samples matched to meta_v2: ", sum(!is.na(meta_idx)), " / ", length(rse_samples)))
if (any(is.na(meta_idx))) {
  log_step(paste0("WARN: ", sum(is.na(meta_idx)), " samples have no meta_v2 entry — colData rows will be NA"))
}

# Build new colData from meta_v2 rows, aligned to rse column order
# Unmatched samples get an all-NA row so dimensions stay intact
new_coldata = meta_v2[meta_idx, , drop = FALSE]
rownames(new_coldata) = colnames(rse)            # must match rse column names

# Drop original colData, attach meta_v2-derived colData
colData(rse) = DataFrame(new_coldata)

log_step(paste0("colData replaced: ", ncol(new_coldata), " metadata columns from meta_v2"))

# --- Subset to protein-coding genes matching TCGA/GTEX ---
idx = match(target_gene_ids, rownames(rse))
log_step(paste0("Genes matched: ", sum(!is.na(idx)), " / ", length(target_gene_ids)))

other_brain_rse_gene = rse[na.omit(idx), ]
rownames(other_brain_rse_gene) = target_gene_ids[!is.na(idx)]

save(other_brain_rse_gene,
     file = file.path(resdir, "other_brain_rse_gene_6_3_2020.RData"))
log_step("Saved other_brain_rse_gene_6_3_2020.RData")
```



# Combining data from recount2

```{r}
# Biologist note:
# This block combines TCGA, GTEx, and CGGA into one harmonized expression matrix,
# then appends glioma-related annotations (IDH, 1p/19q, grade, subtype labels).

library(SummarizedExperiment)
library(rtracklayer)
library(recount)

if (!requireNamespace("futile.logger", quietly = TRUE)) {
  stop("Package 'futile.logger' is required. Please run install.packages('futile.logger').")
}
library(futile.logger)

project_dir = normalizePath(".", winslash = "/", mustWork = FALSE)
project_data_dir = file.path(project_dir, "data")
raw_recount_dir = file.path(project_data_dir, "raw", "recount_brain")
meta_v2 = get(load(file.path(raw_recount_dir, "recount_brain_v2.Rdata")))

data_dir = file.path(project_data_dir, "processed", "log2_tpm_rse")
output_dir = file.path(project_dir, "output")
final_output_dir = file.path(output_dir, "final_processed")
dir.create(final_output_dir, recursive = TRUE, showWarnings = FALSE)

required_files = c(
  file.path(data_dir, "TCGA_rse_gene.RData"),
  file.path(data_dir, "GTEX_rse_gene.RData"),
  file.path(data_dir, "other_brain_rse_gene_6_3_2020.RData")
)
missing_required = required_files[!file.exists(required_files)]
if (length(missing_required) > 0) {
  stop(
    paste0(
      "Missing required processed files in data/processed/log2_tpm_rse:\n- ",
      paste(missing_required, collapse = "\n- "),
      "\nPlease generate/download them before running the combine step."
    )
  )
}

flog.appender(appender.console(), name = "BRAINUMAP")
flog.threshold(INFO, name = "BRAINUMAP")
log_step = function(msg) {
  futile.logger::flog.info(msg, name = "BRAINUMAP")
  message(msg)
}
log_step("START: Combining TCGA, GTEx, and CGGA into harmonized matrix")
tcga_meta = meta_v2[which(meta_v2[,"Dataset"]=="TCGA"), ]
gtex_meta = meta_v2[ which(meta_v2[,"Dataset"]=="GTEX"), ]

# TCGA
tcga_rse = get(load( file.path(data_dir,  "TCGA_rse_gene.RData") ))
idx2 = match(tcga_meta$run_s, colnames(tcga_rse))
table(is.na(idx2))
tcga_rse = tcga_rse[, idx2 ]
if(identical(tcga_meta$run_s, colnames(tcga_rse))){
  rownames(tcga_meta) =tcga_meta$run_s
}
tcga_mat = assay(tcga_rse)

#GTEX
gtex_rse = get(load( file.path(data_dir,  "GTEX_rse_gene.RData") ))
gtex_rse = gtex_rse[, which(colData(gtex_rse)$smts=="Brain")]
idx = match(gtex_meta$run_s, colnames(gtex_rse))
table(is.na(idx))
gtex_rse = gtex_rse[, idx ]
if(identical(gtex_meta$run_s, colnames(gtex_rse))){
  rownames(gtex_meta) =gtex_meta$run_s
}
gtex_mat = assay(gtex_rse)

# CGGA
other_rse = get(load(file.path(data_dir, "other_brain_rse_gene_6_3_2020.RData")))
other_rse$sra_study_s <- "SRP027383"
keep_study =c("SRP027383") # CGGA
keep_col = which(other_rse$sra_study_s %in% keep_study)
other_rse = other_rse[, keep_col]

other_mat = assay(other_rse)
other_meta = as.data.frame(colData(other_rse))

gtex_meta$disease_status = rep("GTEx" , nrow(gtex_meta))
tcga_meta$Dataset2 = as.character (tcga_meta$Dataset)
gtex_meta$Dataset2 = as.character(gtex_meta$Dataset)
other_meta$Dataset2 = "CGGA"

tcga_meta[which(tcga_meta$tissue=="LGG"),"Dataset2"] = "TCGA-LGG"
tcga_meta[which(tcga_meta$tissue=="GBM"),"Dataset2"] = "TCGA-GBM"
tcga_meta$sra_study_s ="TCGA"

# COMBINE MATRIX  & COLDATA FOR COMPUTATION 
mat1 = cbind(tcga_mat, gtex_mat, other_mat)
finaldf = rbind(tcga_meta, gtex_meta, other_meta)


want_cols = c("Dataset2" , "disease" ,
   "age" , "age_units", "development" , "sex", "race", 
   "tissue_site_1" , 
  "clinical_stage_2", # primary, recurrent, solid tumor and so on..
  "viability", # biopsy or post mortem
   "bioproject_s", "run_s",   "sample_name_s" , "sra_sample_s"   ,"sra_study_s", 
  "pmi", "pmi_units")

finaldf = finaldf[, match(want_cols, colnames(finaldf))]

control_idx2 = which(is.na(finaldf$tissue_site_1))
finaldf$tissue_site_1[control_idx2] = "Not available"

finaldf$development[which(is.na(finaldf$development))] = "Not available"

finaldf$tissue_site_1[grep("BA24",  finaldf$tissue_site_1)] = "BA24"
finaldf$tissue_site_1 = gsub("Cerebellar Hemisphere", "Cerebellum", finaldf$tissue_site_1)
finaldf$tissue_site_1[grep("Spinal cord", finaldf$tissue_site_1)]="Spinal cord"


table(finaldf$Dataset2, useNA="always")
table(finaldf$disease, useNA="always")
table(finaldf$development, useNA="always")
table(finaldf$tissue_site_1, useNA="always")






```
```{r}
library(readxl)

cgga_annot_path = "K:/PHO/AG Kerl/Victor/01_PBTA/00_Github/BRAINUMAP/data/raw/annotations/PMC4754110_s1_gbm_lgg_codel_status.xlsx"

# Check sheet names first
readxl::excel_sheets(cgga_annot_path)

# Read first sheet (adjust sheet= if needed)
tcga_codel = readxl::read_excel(cgga_annot_path, sheet = 1)


```


```{r}
# fix - metadata  for Gliomas
# add IDH status for TCGA : 

tcga_codel = tcga_codel[,c("Case","Study", "IDH status", "1p/19q codeletion")]
colnames(tcga_codel) = c("TCGA_case", "Study", "TCGA_idh", "TCGA_1p_19q_codel")
tcga_codel[,3] = gsub("WT", "Wildtype", tcga_codel[,3])
test = merge(x =finaldf, y =tcga_codel, by.x ="bioproject_s", by.y = "TCGA_case", all.x=TRUE)

finaldf = test[match(rownames(finaldf), test$run_s), ]

colnames(finaldf)[match("TCGA_idh", colnames(finaldf))] ="TCGA_Glioma_IDH"
colnames(finaldf)[match("TCGA_1p_19q_codel", colnames(finaldf))] ="TCGA_Glioma_1p_19q_codel"

# add grade for TCGA : 
finaldf$TCGA_Glioma_Grade = rep("Not Available", nrow(finaldf))
finaldf$TCGA_Glioma_Grade[which(finaldf$disease=="TCGA-GBM")] = "4"
finaldf$TCGA_Glioma_Grade[which(finaldf$disease=="TCGA-LGG" &
                             finaldf$clinical_stage_1=="Grade II")] = "2"
finaldf$TCGA_Glioma_Grade[which(finaldf$disease=="TCGA-LGG" &
                             finaldf$clinical_stage_1=="Grade III")] = "3"



```

# CCGA
```{r}

library(futile.logger)

project_dir      = normalizePath(".", winslash = "/", mustWork = FALSE)
project_data_dir = file.path(project_dir, "data")
raw_cgga_dir     = file.path(project_data_dir, "raw", "cgga")
output_dir       = file.path(project_data_dir, "processed")
dir.create(raw_cgga_dir, recursive = TRUE, showWarnings = FALSE)

log_step = function(msg) { futile.logger::flog.info(msg); message(msg) }

# --- Download both CGGA clinical files ---
files = list(
  list(
    url     = "https://www.cgga.org.cn/download?file=download/20200506/CGGA.mRNAseq_693_clinical.20200506.txt.zip",
    zip     = file.path(raw_cgga_dir, "CGGA_693_clinical.zip"),
    pattern = "693_clinical",
    batch   = "mRNAseq_693"
  ),
  list(
    url     = "https://www.cgga.org.cn/download?file=download/20200506/CGGA.mRNAseq_325_clinical.20200506.txt.zip",
    zip     = file.path(raw_cgga_dir, "CGGA_325_clinical.zip"),
    pattern = "325_clinical",
    batch   = "mRNAseq_325"
  )
)


```

```{r}
clin_list = lapply(files, function(f) {

  txt = list.files(raw_cgga_dir, pattern = f$pattern, full.names = TRUE, recursive = TRUE)
  txt = txt[grepl("\\.txt$", txt)][1]
  log_step(paste0("Reading: ", txt))
  df = read.table(txt, header = TRUE, sep = "\t", quote = "", 
                  stringsAsFactors = FALSE, check.names = FALSE)
  df$batch = f$batch
  df
})
```


```{r}
clin_list = lapply(files, function(f) {

  txt = list.files(raw_cgga_dir, pattern = f$pattern, full.names = TRUE, recursive = TRUE)
  txt = txt[grepl("\\.txt$", txt)][1]
  log_step(paste0("Reading: ", txt))
  df = read.table(txt, header = TRUE, sep = "\t", quote = "", 
                  stringsAsFactors = FALSE, check.names = FALSE)
  df$batch = f$batch
  df
})

# --- Find common columns and merge ---
common_cols = Reduce(intersect, lapply(clin_list, colnames))
log_step(paste0("Common columns across batches: ", length(common_cols)))
print(common_cols)

# Merge on common columns (keeps batch column)
cgga_combined = do.call(rbind, lapply(clin_list, function(df) df[, c(common_cols, "batch")]))

log_step(paste0("Combined dimensions: ", nrow(cgga_combined), " samples x ", ncol(cgga_combined), " columns"))

# --- Save ---
out_path = file.path(raw_cgga_dir, "final_cgga_patient_information.txt")
write.table(cgga_combined, file = out_path, sep = "\t", 
            quote = FALSE, row.names = FALSE)
log_step(paste0("Saved: ", out_path))
```


```{r}
# CGGA : 
cgga_key = read.delim(out_path, 
                      header=T, stringsAsFactors = FALSE)
rm_idx = which(is.na( match( cgga_key[,1], finaldf$sample_name_s) ))
cgga_key = cgga_key[-c(rm_idx), ]

length(which(finaldf$sra_study_s=="SRP027383"))
table(is.na( match(finaldf$sample_name_s, cgga_key[,1])))

finaldf$CGGA_Glioma_Grade = rep("Not Available", nrow(finaldf))
finaldf$CGGA_Glioma_1p_19q_codel = rep("Not Available", nrow(finaldf))
finaldf$CGGA_Glioma_IDH  = rep("Not Available", nrow(finaldf))

finaldf$CGGA_Glioma_1p_19q_codel[ match( cgga_key[,1], finaldf$sample_name_s)] = cgga_key$X1p19q_codeletion_status
finaldf$CGGA_Glioma_Grade[ match( cgga_key[,1], finaldf$sample_name_s)] = cgga_key$Grade
finaldf$CGGA_Glioma_IDH[ match( cgga_key[,1], finaldf$sample_name_s)] = cgga_key$IDH_mutation_status

finaldf$CGGA_Glioma_Grade = factor(finaldf$CGGA_Glioma_Grade, levels = c("Not Available", "2", "3", "4"))
finaldf$CGGA_Glioma_IDH = factor(finaldf$CGGA_Glioma_IDH, levels = c("Not Available", "Mutant", "Wildtype"))
finaldf$CGGA_Glioma_1p_19q_codel = factor(finaldf$CGGA_Glioma_1p_19q_codel, levels = c("Not Available", "Codel", "Non-codel"))

# add a column for glioma subtype
finaldf$glioma_subtype = "Not available"
finaldf$glioma_subtype[which(finaldf$TCGA_Glioma_IDH=="Mutant") ] ="TCGA-IDH-Mutant"
finaldf$glioma_subtype[which(finaldf$CGGA_Glioma_IDH=="Mutant") ] ="CGGA-IDH-Mutant"
finaldf$glioma_subtype[which(finaldf$CGGA_Glioma_IDH=="Wildtype") ] ="CGGA-IDH-WT"
finaldf$glioma_subtype[which(finaldf$TCGA_Glioma_IDH=="Wildtype") ] ="TCGA-IDH-WT"
finaldf$glioma_subtype[which(finaldf$CGGA_Glioma_1p_19q_codel=="Codel") ] ="CGGA-Oligo"
finaldf$glioma_subtype[which(finaldf$TCGA_Glioma_1p_19q_codel=="codel") ] ="TCGA-Oligo"
finaldf$glioma_subtype[which(finaldf$clinical_stage_2=="Solid Tissue Normal") ] ="TCGA-GBM-Matched-Normal"
as.data.frame(table(finaldf$glioma_subtype, useNA = "always") )

# Create a Summarized Experiment object
big_rse <- SummarizedExperiment(assays=SimpleList(counts=mat1),
                                rowRanges=rowRanges(tcga_rse), 
                                colData=finaldf)

identical(colnames(big_rse), colData(big_rse)$run_s ) 

# save the files. 
write.table(finaldf, file.path(final_output_dir, "FULL_pca_tsne_umap_combined_analysis_1_28_2021.txt"), 
            sep ="\t", quote=FALSE, row.names=FALSE, col.names=TRUE)
save(big_rse, file =file.path(final_output_dir, "log2_tpm_big_rse_1_28_2021.Rdata"))
log_step("END: Saved final harmonized metadata and SummarizedExperiment to output/final_processed")
```


# Pediatric Data from CBTN 

```{r}
# Biologist note:
# This block processes CBTN counts and converts them to log2(TPM+1),
# restricted to the same shared genes as the adult reference matrix.

rm(list=ls())

library(rtracklayer)
library(edgeR)
library(SummarizedExperiment)

if (!requireNamespace("futile.logger", quietly = TRUE)) {
  stop("Package 'futile.logger' is required. Please run install.packages('futile.logger').")
}
library(futile.logger)

project_dir = normalizePath(".", winslash = "/", mustWork = FALSE)
flog.appender(appender.console(), name = "BRAINUMAP")
flog.threshold(INFO, name = "BRAINUMAP")
log_step = function(msg) {
  futile.logger::flog.info(msg, name = "BRAINUMAP")
  message(msg)
}
log_step("START: Processing CBTN counts and matching to shared genes")

openpbta_dir = file.path(normalizePath(".", winslash = "/", mustWork = FALSE), "data", "raw", "openpbta")
d1 = file.path(openpbta_dir, "release-v23-20230115")
dir.create(d1, recursive = TRUE, showWarnings = FALSE)

cbtn_rds_path = file.path(d1, "pbta-gene-counts-rsem-expected_count.stranded.rds")
if (!file.exists(cbtn_rds_path)) {
  # Optional: set this URL if you have a direct downloadable source for the exact release file.
  pbta_counts_url = Sys.getenv("PBTA_COUNTS_URL", unset = "")
  if (nzchar(pbta_counts_url)) {
    log_step("CBTN counts file missing locally. Attempting download from PBTA_COUNTS_URL.")
    downloader::download(pbta_counts_url, destfile = cbtn_rds_path, mode = "wb")
  }
}

if (!file.exists(cbtn_rds_path)) {
  stop(
    paste0(
      "Missing required input file: ", cbtn_rds_path, "\n",
      "Place pbta-gene-counts-rsem-expected_count.stranded.rds in that folder, ",
      "or set environment variable PBTA_COUNTS_URL to a direct download URL and rerun."
    )
  )
}

cbtn = readRDS(cbtn_rds_path)
temp = cbtn[,1] # first column looks like this : ENSG00000000003.14_TSPAN6
gene_id = sapply(strsplit(temp, "[_]"), function(x) x[1])
length(gene_id)
length(unique(gene_id))
want = unique(gene_id)
midx = match(want, gene_id)
cbtn = cbtn[midx, ]
rownames(cbtn) = want
cbtn = cbtn[,-1]

dim(cbtn) # 58302   977


gtf_v27_path = file.path(openpbta_dir, "gencode.v27.primary_assembly.annotation.gtf.gz")
if (!file.exists(gtf_v27_path)) {
  log_step("Missing gencode.v27.primary_assembly.annotation.gtf.gz - downloading to data/raw/openpbta")
  downloader::download(
    url = "https://ftp.ebi.ac.uk/pub/databases/gencode/Gencode_human/release_27/gencode.v27.primary_assembly.annotation.gtf.gz",
    destfile = gtf_v27_path,
    mode = "wb"
  )
}

gtf = import(gtf_v27_path)
gtf = gtf[which(gtf$type=="gene") , ]
cs = intersect( rownames(cbtn), gtf$gene_id)
length(cs)
cbtn = cbtn[cs, ]
gtf = gtf[match(cs, gtf$gene_id ), ]
 
identical(gtf$gene_id, rownames(cbtn)) # TRUE

# KEEP same genes as my brain-umap

project_dir = normalizePath(".", winslash = "/", mustWork = FALSE)
final_output_dir = file.path(project_dir, "output", "final_processed")
dir.create(final_output_dir, recursive = TRUE, showWarnings = FALSE)
big_rse = get(load(file.path(final_output_dir, "log2_tpm_big_rse_1_28_2021.Rdata")))
big_mat = assay(big_rse)

common_genes= intersect( rowRanges(big_rse)$gencode_gene_name,gtf$gene_name)

midx1 = match(common_genes, gtf$gene_name)
cbtn = cbtn[midx1, ]
rownames(cbtn) =common_genes

midx2 = match(common_genes, rowRanges(big_rse)$gencode_gene_name)
big_rse = big_rse[midx2, ]
big_mat = big_mat[midx2, ]
rownames(big_mat) = common_genes

midx3 = match(common_genes, gtf$gene_name)
gtf = gtf[ midx3, ]

# finally make TPM counts. 
rpkm_data = rpkm(cbtn, gene.length = width(gtf))
tpm_data =  apply(rpkm_data, 2, function(x){
   (x/sum(x))*1000000
})
log2_tpm_cbtn  = log2(tpm_data +1)

saveRDS(log2_tpm_cbtn, file.path(final_output_dir, "log2_tpm_cbtn.rds") )
log_step("END: Saved CBTN log2(TPM+1) matrix to output/final_processed/log2_tpm_cbtn.rds")

```
