---
title: "Process and Harmonize Metadata"
author: "Victor Banerjee"
date: "March 18, 2026"
version: "0.20"
output:
	html_document:
		toc: true
		theme: united
---

## Purpose

This notebook does three things for the three core processed objects:
1. Load TCGA, GTEX, and other_brain SummarizedExperiment objects.
2. Check `colData` for missing values and schema differences.
3. Create a shared `colData` structure (same columns, same order) and save harmonized outputs.

```{r setup}


library(SummarizedExperiment)
library(S4Vectors)

project_dir = normalizePath(".", winslash = "/", mustWork = FALSE)
data_dir = file.path(project_dir, "data", "processed", "log2_tpm_rse")
out_dir = file.path(project_dir, "output", "metadata_checks")
dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)

tcga_path  = file.path(data_dir, "TCGA_rse_gene.RData")
gtex_path  = file.path(data_dir, "GTEX_rse_gene.RData")
other_path = file.path(data_dir, "other_brain_rse_gene_6_3_2020.RData")

needed = c(tcga_path, gtex_path, other_path)
missing_needed = needed[!file.exists(needed)]
if (length(missing_needed) > 0) {
	stop(
		paste0(
			"Missing required files:\n- ",
			paste(missing_needed, collapse = "\n- ")
		)
	)
}
```

## Load objects

```{r load_objects}
tcga_rse  = get(load(tcga_path))
gtex_rse  = get(load(gtex_path))
other_rse = get(load(other_path))

obj_list = list(
	TCGA = tcga_rse,
	GTEX = gtex_rse,
	OTHER = other_rse
)

print(sapply(obj_list, ncol))
```



# GTEX data
### https://www.science.org/doi/10.1126/science.aaz1776



```{r}


# File 2: Subject Phenotypes (sex, age range, Hardy scale — one row per donor)
subject_phe_url  = "https://storage.googleapis.com/adult-gtex/annotations/v8/metadata-files/GTEx_Analysis_v8_Annotations_SubjectPhenotypesDS.txt"
subject_phe_path = file.path(gtex_anno_dir, "GTEx_Analysis_v8_Annotations_SubjectPhenotypesDS.txt")

if (!file.exists(subject_phe_path)) {
  log_step("Downloading GTEx v8 SubjectPhenotypesDS")
  download.file(subject_phe_url, destfile = subject_phe_path, mode = "wb")
}

gtex_subjects = read.delim(subject_phe_path, stringsAsFactors = FALSE, check.names = FALSE)

gtex_subjects = gtex_subjects %>%
  mutate(
    SEX = case_when(
      SEX == 1 ~ "Male",
      SEX == 2 ~ "Female",
      TRUE     ~ NA_character_
    ),
    DTHHRDY = case_when(
      DTHHRDY == 0 ~ "Ventilator",
      DTHHRDY == 1 ~ "Violent/fast",
      DTHHRDY == 2 ~ "Fast natural",
      DTHHRDY == 3 ~ "Intermediate",
      DTHHRDY == 4 ~ "Slow death",
      TRUE         ~ NA_character_
    ),
    # AGE is already a range string e.g. "60-69" — just rename for clarity
    age_range = AGE
  ) %>%
  rename(
    donor_id       = SUBJID,
    sex            = SEX,
    death_classification = DTHHRDY
  ) %>%
  select(donor_id, sex, age_range, death_classification)

gtex_subjects
```





```{r}
library(dplyr)

brain_df = colData(gtex_rse) %>%
  as.data.frame() %>%
  mutate(.idx = row_number()) %>%
  filter(smts == "Brain") %>%
  mutate(donor_id = sub("^(GTEX-[^-]+)-.*", "\\1", sampid)) %>%
  left_join(
    gtex_subjects %>% select(donor_id, sex, age_range, death_classification),
    by = "donor_id"
  )

brain_df %>% select(sample, smts , sample,donor_id,sex, sampid,age_range,death_classification, bigwig_file,  smtsd, project, smtstptref)
```
```{r}
col_meta_gtex = brain_df %>%
  mutate(
    # --- Identity ---
    sample_id               = sampid,
    dataset                 = "GTEx",
    source_study            = project,

    # --- Demographics ---
    # sex, age_range, death_classification joined from gtex_subjects

    # --- Tissue ---
    tissue                  = "Normal Brain",
    tumor_type              = "Normal Brain",
    tissue_site_broad       = smts,
    tissue_site_detail      = smtsd,
    brain_region            = smtsd,

    # --- Sample quality ---
    rna_integrity_number    = as.character(smrin),
    postmortem_interval_min = as.character(smtspax),
    autolysis_score         = as.character(smatsscr),

    # --- SRA traceability ---
    sra_run                 = run,
    sra_sample              = sample,
    sra_experiment          = experiment,

    sample_type             = "Normal"
  ) %>%
  select(
    sample_id, dataset, source_study,
    donor_id, sex, age_range, death_classification,
    tissue, tumor_type,
    tissue_site_broad, tissue_site_detail, brain_region,
    rna_integrity_number, postmortem_interval_min, autolysis_score,
    sra_run, sra_sample, sra_experiment,
    sample_type
  )

# Attach to RSE
rownames(col_meta_gtex) = colnames(brain_gtex_rse)
colData(brain_gtex_rse) = DataFrame(col_meta_gtex)

log_step(paste0("GTEx colData harmonized: ", nrow(col_meta_gtex), " samples x ",
                ncol(col_meta_gtex), " columns"))
log_step(paste0("  Brain regions: ",
                paste(sort(unique(col_meta_gtex$tissue_site_detail)), collapse = " | ")))
log_step(paste0("  Sex coverage:  ",
                sum(!is.na(col_meta_gtex$sex)), " / ", nrow(col_meta_gtex)))
log_step(paste0("  Age coverage:  ",
                sum(!is.na(col_meta_gtex$age_range)), " / ", nrow(col_meta_gtex)))
log_step(paste0("  Hardy scale:   ",
                sum(!is.na(col_meta_gtex$death_classification)), " / ", nrow(col_meta_gtex)))
```

```{r}
as.data.frame(colData(brain_gtex_rse) )
```


```{r}
brain_gtex_rse <-  colData(gtex_rse) %>%
  as.data.frame() %>%
  filter(smts == "Brain") 

brain_gtex_rse
```

```{r}
saveRDS(brain_gtex_rse, file.path(out_dir, "GTExBrain.Final.SEObject.rds")) 
```

# CBTN
```{r}
final_output_dir = file.path(project_dir, "output", "final_processed")  

cbtn_rse <- readRDS(file.path(final_output_dir, "cbtn_pedrictic.rds"))
```


```{r}
cbtn_df <- as.data.frame(colData(cbtn_rse))
cbtn_df
```

```{r}
cbtn_df %>% mutate(
  # --- Identity ---
  sample_id          = Kids_First_Biospecimen_ID,
  dataset            = "CBTN",                          # "CBTN"
  source_study       = cohort,

  # --- Demographics ---
  sex                = reported_gender,
  age_range          = age_at_diagnosis_days,           # days; rename for clarity

  # --- Tissue ---
  tissue             = "Tumor Brain",
  brain_region       = CNS_region,
  histology          = broad_histology,
  who_grade          = NA_character_,                   # not directly available; derive from pathology_diagnosis if needed
  tumor_type         = cancer_group,

  # --- Molecular subtype ---
  idh_status         = NA_character_,                   # not in CBTN metadata; add if merged from molecular data
  codel_1p19q        = NA_character_,
  molecular_subtype  = molecular_subtype,
  mgmt_status        = NA_character_,
  tert_status        = NA_character_,

  # --- Survival ---
  os_days            = as.character(OS_days),
  os_status          = OS_status,
  vital_status       = OS_status,                       # CBTN uses OS_status as proxy
  pfs_days           = as.character(PFS_days),
  pfs_status         = NA_character_,                   # not explicitly in CBTN metadata
  dfs_days           = NA_character_,
  dfs_status         = NA_character_,
  days_to_last_followup = as.character(age_last_update_days),
  sample_type        = sample_type,
  treatment          = extent_of_tumor_resection,       # closest surgical/treatment proxy


  # --- Tissue site ---
  tissue_site_broad  = primary_site,
  tissue_site_detail = primary_site,

  # --- CBTN-specific ---
  biospecimen_id     = sample_id,
  participant_id     = Kids_First_Participant_ID,
  cancer_predisposition = cancer_predispositions,
  tumor_descriptor   = tumor_descriptor
  
) %>% select(
  sample_id, dataset, source_study,
  donor_id,                                             # add cohort_participant_id as donor_id before this select
  sex, age_range,
  tissue, brain_region, histology, who_grade, tumor_type,
  idh_status, codel_1p19q, molecular_subtype, mgmt_status, tert_status,
  os_days, os_status, vital_status,
  pfs_days, pfs_status, dfs_days, dfs_status, days_to_last_followup,
  sample_type, treatment,
  rna_integrity_number, postmortem_interval_min, autolysis_score, death_classification,
  tissue_site_broad, tissue_site_detail,
  biospecimen_id, participant_id, cancer_predisposition, tumor_descriptor,
  sra_run, sra_sample, sra_experiment
)

```



## Check colData before harmonization


```{r check_coldata_before}
colnames_list = lapply(obj_list, function(x) colnames(as.data.frame(colData(x))))

all_cols_union = unique(unlist(colnames_list))
cat("Total unique metadata columns across all 3 objects:", length(all_cols_union), "\n")

schema_summary = data.frame(
	column = all_cols_union,
	in_TCGA = all_cols_union %in% colnames_list$TCGA,
	in_GTEX = all_cols_union %in% colnames_list$GTEX,
	in_OTHER = all_cols_union %in% colnames_list$OTHER,
	stringsAsFactors = FALSE
)

write.table(
	schema_summary,
	file = file.path(out_dir, "coldata_schema_presence_before.tsv"),
	sep = "\t",
	quote = FALSE,
	row.names = FALSE
)

na_summary_before = do.call(
	rbind,
	lapply(names(obj_list), function(nm) {
		cd = as.data.frame(colData(obj_list[[nm]]), stringsAsFactors = FALSE)
		data.frame(
			object = nm,
			column = names(cd),
			n_na = sapply(cd, function(v) sum(is.na(v))),
			pct_na = round(100 * sapply(cd, function(v) mean(is.na(v))), 2),
			stringsAsFactors = FALSE
		)
	})
)

write.table(
	na_summary_before,
	file = file.path(out_dir, "coldata_na_summary_before.tsv"),
	sep = "\t",
	quote = FALSE,
	row.names = FALSE
)

head(na_summary_before)
```

## Harmonize colData to a shared schema

```{r harmonize_coldata}
# Put biologically important columns first (if present), then all others.
priority_cols = c(
	"run_s", "Dataset", "Dataset2", "disease", "disease_status", "sample_name_s",
	"sra_sample_s", "sra_study_s", "tissue", "sex", "age", "development",
	"tissue_site_1", "clinical_stage_1", "clinical_stage_2", "bioproject_s"
)

shared_cols = unique(c(priority_cols, all_cols_union))

harmonize_one_coldata = function(rse_obj, shared_cols, object_name) {
	cd = as.data.frame(colData(rse_obj), stringsAsFactors = FALSE)

	# Add missing columns as NA_character_ so all objects have identical metadata structure.
	missing_cols = setdiff(shared_cols, colnames(cd))
	if (length(missing_cols) > 0) {
		for (m in missing_cols) {
			cd[[m]] = NA_character_
		}
	}

	# Keep same column order everywhere.
	cd = cd[, shared_cols, drop = FALSE]

	# Ensure run_s exists and is aligned to sample columns.
	if (!("run_s" %in% colnames(cd))) {
		cd$run_s = colnames(rse_obj)
	}
	bad_run = is.na(cd$run_s) | cd$run_s == ""
	if (any(bad_run)) {
		cd$run_s[bad_run] = colnames(rse_obj)[bad_run]
	}

	# rownames in colData must match object columns.
	rownames(cd) = colnames(rse_obj)

	colData(rse_obj) = S4Vectors::DataFrame(cd)

	cat(object_name, "-> colData columns:", ncol(cd), "| rows:", nrow(cd), "\n")
	return(rse_obj)
}

tcga_rse_h  = harmonize_one_coldata(tcga_rse,  shared_cols, "TCGA")
gtex_rse_h  = harmonize_one_coldata(gtex_rse,  shared_cols, "GTEX")
other_rse_h = harmonize_one_coldata(other_rse, shared_cols, "OTHER")
```

## Validate and save harmonized objects

```{r validate_and_save}
stopifnot(identical(colnames(colData(tcga_rse_h)), colnames(colData(gtex_rse_h))))
stopifnot(identical(colnames(colData(tcga_rse_h)), colnames(colData(other_rse_h))))

na_summary_after = do.call(
	rbind,
	lapply(
		list(TCGA = tcga_rse_h, GTEX = gtex_rse_h, OTHER = other_rse_h),
		function(x) {
			cd = as.data.frame(colData(x), stringsAsFactors = FALSE)
			data.frame(
				object = unique(ifelse(grepl("TCGA", colnames(cd)[1]), "TCGA", NA)),
				column = names(cd),
				n_na = sapply(cd, function(v) sum(is.na(v))),
				pct_na = round(100 * sapply(cd, function(v) mean(is.na(v))), 2),
				stringsAsFactors = FALSE
			)
		}
	)
)

# Fix object labels for after-summary
na_summary_after$object = rep(c("TCGA", "GTEX", "OTHER"), each = ncol(colData(tcga_rse_h)))

write.table(
	na_summary_after,
	file = file.path(out_dir, "coldata_na_summary_after.tsv"),
	sep = "\t",
	quote = FALSE,
	row.names = FALSE
)

# Save new harmonized versions (original files are preserved).
save(tcga_rse_h,  file = file.path(data_dir, "TCGA_rse_gene_coldata_harmonized.RData"))
save(gtex_rse_h,  file = file.path(data_dir, "GTEX_rse_gene_coldata_harmonized.RData"))
save(other_rse_h, file = file.path(data_dir, "other_brain_rse_gene_6_3_2020_coldata_harmonized.RData"))

# Optional combined metadata table for quick inspection.
combined_coldata = rbind(
	as.data.frame(colData(tcga_rse_h), stringsAsFactors = FALSE),
	as.data.frame(colData(gtex_rse_h), stringsAsFactors = FALSE),
	as.data.frame(colData(other_rse_h), stringsAsFactors = FALSE)
)

write.table(
	combined_coldata,
	file = file.path(out_dir, "combined_coldata_harmonized.tsv"),
	sep = "\t",
	quote = FALSE,
	row.names = FALSE
)

cat("Harmonization complete.\n")
cat("Saved harmonized objects and metadata checks under:\n", out_dir, "\n")
```

