---
title: "Process and Harmonize Metadata"
author: "Victor Banerjee"
date: "March 18, 2026"
version: "0.10"
output:
	html_document:
		toc: true
		theme: united
---

## Purpose

This notebook does three things for the three core processed objects:
1. Load TCGA, GTEX, and other_brain SummarizedExperiment objects.
2. Check `colData` for missing values and schema differences.
3. Create a shared `colData` structure (same columns, same order) and save harmonized outputs.

```{r setup}
rm(list = ls())

library(SummarizedExperiment)
library(S4Vectors)

project_dir = normalizePath(".", winslash = "/", mustWork = FALSE)
data_dir = file.path(project_dir, "data", "processed", "log2_tpm_rse")
out_dir = file.path(project_dir, "output", "metadata_checks")
dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)

tcga_path  = file.path(data_dir, "TCGA_rse_gene.RData")
gtex_path  = file.path(data_dir, "GTEX_rse_gene.RData")
other_path = file.path(data_dir, "other_brain_rse_gene_6_3_2020.RData")

needed = c(tcga_path, gtex_path, other_path)
missing_needed = needed[!file.exists(needed)]
if (length(missing_needed) > 0) {
	stop(
		paste0(
			"Missing required files:\n- ",
			paste(missing_needed, collapse = "\n- ")
		)
	)
}
```

## Load objects

```{r load_objects}
tcga_rse  = get(load(tcga_path))
gtex_rse  = get(load(gtex_path))
other_rse = get(load(other_path))

obj_list = list(
	TCGA = tcga_rse,
	GTEX = gtex_rse,
	OTHER = other_rse
)

print(sapply(obj_list, ncol))
```

## Check colData before harmonization

```{r check_coldata_before}
colnames_list = lapply(obj_list, function(x) colnames(as.data.frame(colData(x))))

all_cols_union = unique(unlist(colnames_list))
cat("Total unique metadata columns across all 3 objects:", length(all_cols_union), "\n")

schema_summary = data.frame(
	column = all_cols_union,
	in_TCGA = all_cols_union %in% colnames_list$TCGA,
	in_GTEX = all_cols_union %in% colnames_list$GTEX,
	in_OTHER = all_cols_union %in% colnames_list$OTHER,
	stringsAsFactors = FALSE
)

write.table(
	schema_summary,
	file = file.path(out_dir, "coldata_schema_presence_before.tsv"),
	sep = "\t",
	quote = FALSE,
	row.names = FALSE
)

na_summary_before = do.call(
	rbind,
	lapply(names(obj_list), function(nm) {
		cd = as.data.frame(colData(obj_list[[nm]]), stringsAsFactors = FALSE)
		data.frame(
			object = nm,
			column = names(cd),
			n_na = sapply(cd, function(v) sum(is.na(v))),
			pct_na = round(100 * sapply(cd, function(v) mean(is.na(v))), 2),
			stringsAsFactors = FALSE
		)
	})
)

write.table(
	na_summary_before,
	file = file.path(out_dir, "coldata_na_summary_before.tsv"),
	sep = "\t",
	quote = FALSE,
	row.names = FALSE
)

head(na_summary_before)
```

## Harmonize colData to a shared schema

```{r harmonize_coldata}
# Put biologically important columns first (if present), then all others.
priority_cols = c(
	"run_s", "Dataset", "Dataset2", "disease", "disease_status", "sample_name_s",
	"sra_sample_s", "sra_study_s", "tissue", "sex", "age", "development",
	"tissue_site_1", "clinical_stage_1", "clinical_stage_2", "bioproject_s"
)

shared_cols = unique(c(priority_cols, all_cols_union))

harmonize_one_coldata = function(rse_obj, shared_cols, object_name) {
	cd = as.data.frame(colData(rse_obj), stringsAsFactors = FALSE)

	# Add missing columns as NA_character_ so all objects have identical metadata structure.
	missing_cols = setdiff(shared_cols, colnames(cd))
	if (length(missing_cols) > 0) {
		for (m in missing_cols) {
			cd[[m]] = NA_character_
		}
	}

	# Keep same column order everywhere.
	cd = cd[, shared_cols, drop = FALSE]

	# Ensure run_s exists and is aligned to sample columns.
	if (!("run_s" %in% colnames(cd))) {
		cd$run_s = colnames(rse_obj)
	}
	bad_run = is.na(cd$run_s) | cd$run_s == ""
	if (any(bad_run)) {
		cd$run_s[bad_run] = colnames(rse_obj)[bad_run]
	}

	# rownames in colData must match object columns.
	rownames(cd) = colnames(rse_obj)

	colData(rse_obj) = S4Vectors::DataFrame(cd)

	cat(object_name, "-> colData columns:", ncol(cd), "| rows:", nrow(cd), "\n")
	return(rse_obj)
}

tcga_rse_h  = harmonize_one_coldata(tcga_rse,  shared_cols, "TCGA")
gtex_rse_h  = harmonize_one_coldata(gtex_rse,  shared_cols, "GTEX")
other_rse_h = harmonize_one_coldata(other_rse, shared_cols, "OTHER")
```

## Validate and save harmonized objects

```{r validate_and_save}
stopifnot(identical(colnames(colData(tcga_rse_h)), colnames(colData(gtex_rse_h))))
stopifnot(identical(colnames(colData(tcga_rse_h)), colnames(colData(other_rse_h))))

na_summary_after = do.call(
	rbind,
	lapply(
		list(TCGA = tcga_rse_h, GTEX = gtex_rse_h, OTHER = other_rse_h),
		function(x) {
			cd = as.data.frame(colData(x), stringsAsFactors = FALSE)
			data.frame(
				object = unique(ifelse(grepl("TCGA", colnames(cd)[1]), "TCGA", NA)),
				column = names(cd),
				n_na = sapply(cd, function(v) sum(is.na(v))),
				pct_na = round(100 * sapply(cd, function(v) mean(is.na(v))), 2),
				stringsAsFactors = FALSE
			)
		}
	)
)

# Fix object labels for after-summary
na_summary_after$object = rep(c("TCGA", "GTEX", "OTHER"), each = ncol(colData(tcga_rse_h)))

write.table(
	na_summary_after,
	file = file.path(out_dir, "coldata_na_summary_after.tsv"),
	sep = "\t",
	quote = FALSE,
	row.names = FALSE
)

# Save new harmonized versions (original files are preserved).
save(tcga_rse_h,  file = file.path(data_dir, "TCGA_rse_gene_coldata_harmonized.RData"))
save(gtex_rse_h,  file = file.path(data_dir, "GTEX_rse_gene_coldata_harmonized.RData"))
save(other_rse_h, file = file.path(data_dir, "other_brain_rse_gene_6_3_2020_coldata_harmonized.RData"))

# Optional combined metadata table for quick inspection.
combined_coldata = rbind(
	as.data.frame(colData(tcga_rse_h), stringsAsFactors = FALSE),
	as.data.frame(colData(gtex_rse_h), stringsAsFactors = FALSE),
	as.data.frame(colData(other_rse_h), stringsAsFactors = FALSE)
)

write.table(
	combined_coldata,
	file = file.path(out_dir, "combined_coldata_harmonized.tsv"),
	sep = "\t",
	quote = FALSE,
	row.names = FALSE
)

cat("Harmonization complete.\n")
cat("Saved harmonized objects and metadata checks under:\n", out_dir, "\n")
```

